Tilting at Algorithmic Windmills

### 2013-05-14 Tilting/Rpaper.Rnw
WIP: adding note from satra
ENH:  Downpressor man.

\author[label1]{Nicholas J.~Tustison
  \fnref{label0}}
  \fntext[label0]{\scriptsize Corresponding author:  PO Box 801339, Charlottesville, VA 22908; T:  434-924-7730; email address:  ntustison@virginia.edu.
  }
\author[label2]{Brian B.~Avants (combined first author, second author or senior author?)}
\address[label1]{Department of Radiology and Medical Imaging, University of Virginia, Charlottesville, VA}
\address[label2]{Penn Image Computing and Science Laboratory, University of Pennsylvania,
                Philadelphia, PA}

@@

As fellow scientists who actively participate in this 
research through our many contributions to the Insight
Toolkit and our own Advanced Normalization Tools (ANTs),%
\footnote{
http://www.picsl.upenn.edu/ANTs
} 
we have noticed several publications in which a principle
component involves the comparison of algorithms.  One of our
concerns is the lack of detail with which these comparisons are
presented and the corresponding possibility of {\it instrumentation
bias} \cite{sackett1979} (considering software  as a type of 
instrument requiring proper ``calibration'' for accurate measurements).  
Based on our experience (including mistakes that we have made), 
we propose an initial set of guidelines for minimal
reporting of algorithmic usage to minimize such bias, understanding that 
the discussion will hopefully 
elicit a much more comprehensive response from the larger community.

@@

\subsection{Comparisons should be performed on publicly available data.}`
For reasons of reproducibility and transparency, evaluations should be performed
using publicly availabile data sets.  Given the rather large number
of such institutional efforts including 
IXI,%
\footnote{
http://www.brain-development.org
}
NKI,%
\footnote{
http://fcon\_1000.projects.nitrc.org/indi/pro/nki.html
}
OASIS,%
\footnote{
http://www.oasis-brains.org
}
Kirby,%
\footnote{
http://mri.kennedykrieger.org/databases.html
},
LONI,%
\footnote{
http://www.loni.ucla.edu/Research/Databases/
}, and
others, evaluations should include (if not be exhausted by)
comparisons using such data.  While evaluation on specific abnormal cohorts
might exclude such possibilities, such specific evaluations should
be extensively motivated in the introduction and/or discussion.
For example, if a particular algorithm with general application is 
found to perform better
on a private cohort of Parkinson's disease subject data, reasons
for performance disparity should be offered and supplemented with
analysis on public data.

@@

\subsection{Introduction}
As part of the recent Medical Image Computing and Computer-Aided Intervention 
(MICCAI) 2012 international meeting, several groups participated in the 
Multi-Atlas Labeling challenge%
\footnote{
https://masi.vuse.vanderbilt.edu/workshop2012
}
where participants proposed algorithmic workflows to propagate
labels from expertly annotated brains to testing data.  One of 
the overall conclusions drawn from the meeting was the significance 
of spatial normalization and so the organizers (Profs. Bennet Landman and Simon
Warfield) suggested rerunning the competition using a set
of transforms derived from a single registration algorithm.
Since three of the top four performing algorithms all used
the \verb#ANTS# registration tool \cite{avants2011} (available
in the Advanced Normalization Tools open source 
toolkit%
\footnote{
http://picsl.upenn.edu/ANTS
}) the organizers approached one of the principle authors of
the ANTs package (B.A.) to provide the ``canonical'' 
\verb#ANTS#-derived transforms.  
However, as part of the recent ITKv4 effort, an improved
version for image registration, known as \verb#antsRegistration#
had been created and also released within ANTs \cite{avants2012}.
Motivated by this request and also to provide an example of the 
guidelines previously presented, we performed a comparison of 
 \verb#ANTS# and \verb#antsRegistration#.  The principle hypothesis
 to be explored with relevance to the challenge is:  {\it Does
 \verb#antsRegistration# outperform \verb#ANTS# in propagating
 labels for the MICCAI challenge training data?}  Important 
 corrollary hypotheses involve similar assessments of both the
 affine and deformable (i.e. SyN) components of the respective
 algorithms.

\subsection{Materials and Methods}

@@

\lstset{frame = tb,
        framerule = 0.25pt,
        float,
        fontadjust,
        backgroundcolor={\color{listlightgray}},
        basicstyle = {\ttfamily\scriptsize},
        keywordstyle = {\ttfamily\color{listkeyword}\textbf},
        identifierstyle = {\ttfamily},
        commentstyle = {\ttfamily\color{listcomment}\textit},
        stringstyle = {\ttfamily},
        showstringspaces = false,
        showtabs = false,
        numbers = none,
        numbersep = 6pt,
        numberstyle={\ttfamily\color{listnumbers}},
        tabsize = 2,
        language=,
        floatplacement=!h,
        caption={\small \baselineskip 12pt ANTS command line call used for the evaluation}.,
        captionpos=b,
        label=listing:ANTS
        }
\begin{lstlisting}
${ANTsPATH}/ImageMath 3 $fixedTruncated \
  TruncateImageIntensity $fixed 0.01 0.99
${ANTsPATH}/ImageMath 3 $movingTruncated \
  TruncateImageIntensity $moving 0.01 0.99

${ANTsPATH}/ANTS 3 -o ${outputPrefix} \
  -m CC[${fixedTruncated},${movingTruncated},1,4] \
  -t SyN[0.1] -r Gauss[3,0] -i 100x100x70x20 \
  --use-Histogram-Matching
\end{lstlisting}

\lstset{frame = tb,
        framerule = 0.25pt,
        float,
        fontadjust,
        backgroundcolor={\color{listlightgray}},
        basicstyle = {\ttfamily\scriptsize},
        keywordstyle = {\ttfamily\color{listkeyword}\textbf},
        identifierstyle = {\ttfamily},
        commentstyle = {\ttfamily\color{listcomment}\textit},
        stringstyle = {\ttfamily},
        showstringspaces = false,
        showtabs = false,
        numbers = none,
        numbersep = 6pt,
        numberstyle={\ttfamily\color{listnumbers}},
        tabsize = 2,
        language=,
        floatplacement=!h,
        caption={\small \baselineskip 12pt antsRegistration command line call used for the evaluation}.,
        captionpos=b,
        label=listing:antsRegistration
        }
\begin{lstlisting}  
${ANTsPATH}/antsRegistration -d 3 -o ${outputPrefix} \
  -u 1 -w [0.01,0.99], -r [${fixed},${moving},1] \
  -t Rigid[0.1] -c [1000x500x250x100,1e-8,10] \
  -s 4x2x1x0 -f 8x4x2x1 \
  -m MI[${fixed},${moving},1,32,Regular,0.25] \
  -t Affine[0.1] -c [1000x500x250x100,1e-8,10] \
  -s 4x2x1x0 -f 8x4x2x1 \
  -m MI[${fixed},${moving},1,32,Regular,0.25] \
  -t SyN[0.1,3,0] -c [100x100x70x20,1e-9,15] \
  -s 3x2x1x0 -f 6x4x2x1 \
  -m CC[${fixed},${moving},1,4]
\end{lstlisting}

@@

\lstset{frame = tb,
        framerule = 0.25pt,
        float,
        fontadjust,
        backgroundcolor={\color{listlightgray}},
        basicstyle = {\ttfamily\scriptsize},
        keywordstyle = {\ttfamily\color{listkeyword}\textbf},
        identifierstyle = {\ttfamily},
        commentstyle = {\ttfamily\color{listcomment}\textit},
        stringstyle = {\ttfamily},
        showstringspaces = false,
        showtabs = false,
        numbers = none,
        numbersep = 6pt,
        numberstyle={\ttfamily\color{listnumbers}},
        tabsize = 2,
        language=,
        floatplacement=!h,
        caption={\small \baselineskip 12pt antsApplyTransforms command line call used to warp
        the moving label images to the fixed image for subsequent label overlap assessment}.,
        captionpos=b,
        label=listing:antsApplyTransforms
        }
\begin{lstlisting}  
${ANTsPATH}/antsApplyTransforms -d 3 \
  -i $movingLabels -r $fixed \
  -n NearestNeighbor \
  -o ${outputPrefix}LabelsWarped.nii.gz \
  -t $outputWarp -t $outputAffine
\end{lstlisting}
Label overlap assessment for each registration was quantified
using the \verb#LabelOverlapMeasures# available through ANTs and 
implemented in ITK\cite{tustison2009}.
It should noted that only the affine transform files are used
to assess affine algorithmic performance. In order to evaluate
the SyN component in isolation from the affine portion, the 
affine transforms from the better-performing algorithm (strictly
in terms of affine performance) were used as input for the other
algorithm which was then rerun for all data sets.

\subsection{Results}

\subsection{Discussion}

@@

\section{Acknowledgments}

>>>>

\author[label1]{Nicholas J.~Tustison
  \fnref{label0}}
  \fntext[label0]{\scriptsize Corresponding author:  PO Box 801339, Charlottesville, VA 22908; T:  434-924-7730; email address:  ntustison@virginia.edu.
  }
\author[label2]{Brian B.~Avants}
\address[label1]{Department of Radiology and Medical Imaging, University of Virginia, Charlottesville, VA}
\address[label2]{Penn Image Computing and Science Laboratory, University of Pennsylvania,
                Philadelphia, PA}

@@

As fellow scientists who actively participate in this 
research through our many contributions to the Insight
Toolkit and our own Advanced Normalization Tools (ANTs),%
\footnote{
http://www.picsl.upenn.edu/ANTs
} 
we have noticed several publications in which a principle
component involves the comparison of algorithms.  One of our
concerns is the lack of detail with which these comparisons are
often presented and the corresponding possibility of {\it instrumentation
bias} \cite{sackett1979} (considering software  as a type of 
instrument requiring proper ``calibration'' for accurate measurements).  
Based on our experience (including mistakes that we have made), 
we propose an initial set of guidelines for minimal
reporting of algorithmic usage to minimize such bias, understanding that 
the discussion will hopefully 
elicit a much more comprehensive response from the larger community.

@@

\subsection{Comparisons should be performed on publicly available data.}
For reasons of reproducibility and transparency, evaluations should be performed
using publicly availabile data sets.  Given the rather large number
of such institutional efforts including 
IXI,%
\footnote{
http://www.brain-development.org
}
NKI,%
\footnote{
http://fcon\_1000.projects.nitrc.org/indi/pro/nki.html
}
OASIS,%
\footnote{
http://www.oasis-brains.org
}
Kirby,%
\footnote{
http://mri.kennedykrieger.org/databases.html
},
LONI,%
\footnote{
http://www.loni.ucla.edu/Research/Databases/
}, and
others, evaluations should include (if not be exhausted by)
comparisons using such data.  While evaluation on specific abnormal cohorts
might exclude such possibilities, such specific evaluations should
be extensively motivated in the introduction and/or discussion.
For example, if a particular algorithm with general application is 
found to perform better
on a private cohort of Parkinson's disease subject data, reasons
for performance disparity should be offered and supplemented with
analysis on public data.

@@

\subsection{ANTS vs. antsRegistration:  Introduction}
As part of the recent Medical Image Computing and Computer-Aided Intervention 
(MICCAI) 2012 international meeting, several groups participated in the 
Multi-Atlas Labeling challenge%
\footnote{
https://masi.vuse.vanderbilt.edu/workshop2012
}
where participants proposed algorithmic workflows to propagate
labels from expertly annotated brains to testing data.  One of 
the overall conclusions drawn from the meeting was the significance 
of spatial normalization and so the organizers (Profs. Bennet Landman and Simon
Warfield) suggested rerunning the competition using a set
of transforms derived from a single registration algorithm.
Since three of the top four performing algorithms all used
the \verb#ANTS# registration tool \cite{avants2011} (available
in the Advanced Normalization Tools open source 
toolkit%
\footnote{
http://picsl.upenn.edu/ANTS
}) the organizers approached one of the principle authors of
the ANTs package (B.A.) to provide the ``canonical'' 
\verb#ANTS#-derived transforms.  
However, as part of the recent ITKv4 effort, a new
program for image registration, known as \verb#antsRegistration#
has been created and also released within ANTs \cite{avants2012}
where it continues to be refined (in contrast to \verb#ANTS# which
has been more or less stable for quite some time).

Motivated by this request and also to provide an example of the 
guidelines previously presented, we performed a comparison of 
 \verb#ANTS# and \verb#antsRegistration#.  The principle hypothesis
 to be explored with relevance to the challenge is:  {\it Does
 \verb#antsRegistration# outperform \verb#ANTS# in propagating
 labels for the MICCAI challenge training data?}  Important 
 corrollary hypotheses involve similar assessments of both the
 affine and deformable (i.e. SyN) components of the respective
 programs.

\subsection{ANTS vs. antsRegistration:  Materials and Methods}

@@

\lstset{frame = tb,
        framerule = 0.25pt,
        float,
        fontadjust,
        backgroundcolor={\color{listlightgray}},
        basicstyle = {\ttfamily\scriptsize},
        keywordstyle = {\ttfamily\color{listkeyword}\textbf},
        identifierstyle = {\ttfamily},
        commentstyle = {\ttfamily\color{listcomment}\textit},
        stringstyle = {\ttfamily},
        showstringspaces = false,
        showtabs = false,
        numbers = none,
        numbersep = 6pt,
        numberstyle={\ttfamily\color{listnumbers}},
        tabsize = 2,
        language=,
        floatplacement=!h,
        caption={\small \baselineskip 12pt ANTS command line call used for the evaluation. The help menu can be invoked by the command `ANTS --help'.},
        captionpos=b,
        label=listing:ANTS
        }
\begin{lstlisting}
${ANTsPATH}/ImageMath 3 $fixedTruncated \
  TruncateImageIntensity $fixed 0.01 0.99
${ANTsPATH}/ImageMath 3 $movingTruncated \
  TruncateImageIntensity $moving 0.01 0.99

${ANTsPATH}/ANTS 3 -o ${outputPrefix} \
  -m CC[${fixedTruncated},${movingTruncated},1,4] \
  -t SyN[0.1] -r Gauss[3,0] -i 100x100x70x20 \
  --number-of-affine-iterations 0 \
  --use-Histogram-Matching
\end{lstlisting}

\lstset{frame = tb,
        framerule = 0.25pt,
        float,
        fontadjust,
        backgroundcolor={\color{listlightgray}},
        basicstyle = {\ttfamily\scriptsize},
        keywordstyle = {\ttfamily\color{listkeyword}\textbf},
        identifierstyle = {\ttfamily},
        commentstyle = {\ttfamily\color{listcomment}\textit},
        stringstyle = {\ttfamily},
        showstringspaces = false,
        showtabs = false,
        numbers = none,
        numbersep = 6pt,
        numberstyle={\ttfamily\color{listnumbers}},
        tabsize = 2,
        language=,
        floatplacement=!h,
        caption={\small \baselineskip 12pt antsRegistration command line call used for the evaluation.  The help menu can be invoked by the command `antsRegistration --help'.},
        captionpos=b,
        label=listing:antsRegistration
        }
\begin{lstlisting}  
${ANTsPATH}/antsRegistration -d 3 -o ${outputPrefix} \
  -u 1 -w [0.01,0.99], -r [${fixed},${moving},1] \
  -t Rigid[0.1] -c [1000x500x250x100,1e-8,10] \
  -s 4x2x1x0 -f 8x4x2x1 \
  -m MI[${fixed},${moving},1,32,Regular,0.25] \
  -t Affine[0.1] -c [1000x500x250x100,1e-8,10] \
  -s 4x2x1x0 -f 8x4x2x1 \
  -m MI[${fixed},${moving},1,32,Regular,0.25] \
  -t SyN[0.1,3,0] -c [100x100x70x20,1e-9,15] \
  -s 3x2x1x0 -f 6x4x2x1 \
  -m CC[${fixed},${moving},1,4]
\end{lstlisting}

@@

\lstset{frame = tb,
        framerule = 0.25pt,
        float,
        fontadjust,
        backgroundcolor={\color{listlightgray}},
        basicstyle = {\ttfamily\scriptsize},
        keywordstyle = {\ttfamily\color{listkeyword}\textbf},
        identifierstyle = {\ttfamily},
        commentstyle = {\ttfamily\color{listcomment}\textit},
        stringstyle = {\ttfamily},
        showstringspaces = false,
        showtabs = false,
        numbers = none,
        numbersep = 6pt,
        numberstyle={\ttfamily\color{listnumbers}},
        tabsize = 2,
        language=,
        floatplacement=!h,
        caption={\small \baselineskip 12pt antsApplyTransforms command line call used to warp
        the moving label images to the fixed image for subsequent label overlap assessment}.,
        captionpos=b,
        label=listing:antsApplyTransforms
        }
\begin{lstlisting}  
${ANTsPATH}/antsApplyTransforms -d 3 \
  -i $movingLabels -r $fixed \
  -n NearestNeighbor \
  -o ${outputPrefix}LabelsWarped.nii.gz \
  -t $outputWarp -t $outputAffine
\end{lstlisting}
Label overlap assessment for each registration was quantified
using the \verb#LabelOverlapMeasures# available through ANTs and 
implemented in ITK \cite{tustison2009}.
It should noted that only the affine transform files are used
to assess affine algorithmic performance. In order to evaluate
the SyN component in isolation from the affine portion, the 
affine transforms from the better-performing algorithm (strictly
in terms of affine performance) were used as input for the other
algorithm which was then rerun for all data sets.

\subsection{ANTS vs. antsRegistration:  Results}

Upon completion of all pairwise registrations, the labels for the
moving image were warped to the fixed image using \verb#antsApplyTransforms#
as described in Listing \ref{listing:antsApplyTransforms} using 
nearest neighbor interpolation.  This was performed 
for both affine and affine + deformable scenarios.  The
corresponding box plots are given in Figure \ref{fig:results}.

\begin{center}
\begin{figure}
  \includegraphics[width=85mm]{tilting.pdf}
  \caption{Combined Dice values for all labels of the evaluation between the two ANTs-based registration programs including affine-only transformations. \textcolor{red}{Perhaps we should also plot Dice but with the white matter and cerebellum labels removed since they're so large (could they be masking more noticeable differences?).}
  }
  \label{fig:results}
\end{figure}
\end{center}

Pairwise Students t-tests were performed for each of the three
comparisons with Bonferroni correction.
For the affine-only Dice value comparison, \verb#antsRegistration# 
only slightly output performed \verb#ANTS# ($q = 0.053$, mean of the 
differences = 0.004).  For the comprehensive affine + SyN comparison
\verb#antsRegistration# yielded higher Dice values ($q = 0.053$, 
mean of the differences = 0.009).
Due to the difference in affine-only outcome, the \verb#ANTS# 
SyN deformable component was run again with initialization from
the affine output of the \verb#antsRegistration# runs.  This
latter set is denoted in Figure \ref{fig:results} as ``ANTS with
other affine.''  This scenario also showed \verb#antsRegistration# SyN 
producing slightly higher values than its ANTs counterpart 
($q = 0.053$, mean of the differences = 0.006).

%\begin{table}
%\begin{tabular*}{0.45\textwidth}{@{\extracolsep{\fill}}lr}
%  Comparison & q-value \\
%  \toprule
%  ANTS vs. antsRegistration (affine only) & 0.053 \\
%  ANTS vs. antsRegistration (affine + SyN) & 2.7e-14 \\
%  ANTS vs. antsRegistration (affine' + SyN) & 3.4e-13 \\
%  \bottomrule
%\end{tabular*}
%\end{table}

\subsection{ANTS vs. antsRegistration:  Discussion}
Strictly in terms of the Dice overlap metric, \verb#antsRegistration# yields
improved results over \verb#ANTS#.  Although the SyN components are theoretically
the same, implementation with the newer program was constrained to conform
with the rigorous ITK coding standards and the plug-in architecture of
the generic registration framework resulting in additional possible
configurations not in \verb#ANTS#. Additionally, the newer program benefitted
from ITK's crowdsourcing of the contributed code (a great example of Linus's Law
in action---"with enough eyeballs, all bugs are shallow").  In contrast to the
SyN component, the affine components between the two programs differ substantially.
The parameters governing the affine registration in \verb#ANTS# were specifically 
tailored for brain registration and much of it is hard-coded internally within the
code.  In contrast, with \verb#antsRegistration# we took a much more general approach
by permitting the user to build up any number of linear ``registration stages'' 
to produce a final generic affine transform.

Other differences to be considered include the use of real types between the
two programs and the use of multi-threading.  As regards the former, due to 
the constraints of certain  internals of ITK,  \verb#antsRegistration#
is currently implemented using \verb#double# instead of \verb#float#.  This
results in a larger memory footprint in addition to timing between the two
programs (single-threaded on a single core on the UVA cluster \verb#antsRegistration#
can run 3-4 times as long as \verb#ANTS#).  However, multi-threading capabilities
are much better developed in \verb#antsRegistration#.

@@

Considering the permeation of the computational sciences in doing 
nueroimaging research, certain safeguards should be in place to prevent
(or at least minimize) potential biases and errors that can
unknowingly effect study outcomes.  There is no vetting agency for
ensuring that analysis programs used for research are reasonably error-free.
In addition, these software packages are simply ``black
boxes'' to many researchers who are not formally trained to debug
code.  And even to those of us who are, debugging other people's code
is oftentimes very difficult.  To this end, algorithmic comparisons are 
a very good way of evaluating general performance.  It is hoped that the
proposed set of guidelines set forth in this editorial aid the community
for future comparative assessments.
